---
title: 'Foundations for statistical inference - Confidence intervals'
output:
  html_document:
    theme: simplex
    highlight: haddock
    css: ..\labs.css
    toc: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(dplyr)
library(ggplot2)
library(mosaic)
set.seed(9438024)

normTail <- function (m = 0, s = 1, L = NULL, U = NULL, M = NULL, df = 1000, 
    curveColor = 1, border = 1, col = "#CCCCCC", xlim = NULL, 
    ylim = NULL, xlab = "", ylab = "", digits = 2, axes = 1, 
    detail = 999, xLab = c("number", "symbol"), cex.axis = 1, 
    xAxisIncr = 1, ...) 
{
    if (is.null(xlim)[1]) {
        xlim <- m + c(-1, 1) * 3.5 * s
    }
    temp <- diff(range(xlim))
    x <- seq(xlim[1] - temp/4, xlim[2] + temp/4, length.out = detail)
    y <- dt((x - m)/s, df)/s
    if (is.null(ylim)[1]) {
        ylim <- range(c(0, y))
    }
    #cat(x)
    #cat(y)
    plot(x, y, type = "l", xlim = xlim, ylim = ylim, xlab = xlab, 
        ylab = ylab, axes = FALSE, col = curveColor, ...)
    if (!is.null(L[1])) {
        these <- (x <= L)
        X <- c(x[these][1], x[these], rev(x[these])[1])
        Y <- c(0, y[these], 0)
        polygon(X, Y, border = border, col = col)
    }
    if (!is.null(U[1])) {
        these <- (x >= U)
        X <- c(x[these][1], x[these], rev(x[these])[1])
        Y <- c(0, y[these], 0)
        polygon(X, Y, border = border, col = col)
    }
    if (all(!is.null(M[1:2]))) {
        these <- (x >= M[1] & x <= M[2])
        X <- c(x[these][1], x[these], rev(x[these])[1])
        Y <- c(0, y[these], 0)
        polygon(X, Y, border = border, col = col)
    }
    if (axes == 1 || axes > 2) {
        if (xLab[1] == "symbol") {
            xAt <- m + (-3:3) * s
            xLab <- expression(mu - 3 * sigma, mu - 2 * sigma, 
                mu - sigma, mu, mu + sigma, mu + 2 * sigma, mu + 
                  3 * sigma)
        }
        else if (xLab[1] != "number") {
            stop("Argument \"xLab\" not recognized.\n")
        }
        else {
            temp <- seq(xAxisIncr, max(abs(xlim - m))/s, xAxisIncr) * 
                s
            xAt <- m + c(-temp, 0, temp)
            xLab <- round(xAt, digits = digits)
        }
    }
    if (axes > 2) {
        axis(1, at = xAt, labels = xLab, cex.axis = cex.axis)
        buildAxis(2, c(y, 0), n = 3, nMax = 3, cex.axis = cex.axis)
    }
    else if (axes > 1) {
        buildAxis(2, c(y, 0), n = 3, nMax = 3, cex.axis = cex.axis)
    }
    else if (axes > 0) {
        axis(1, at = xAt, labels = xLab, cex.axis = cex.axis)
    }
    abline(h = 0)
}
```

************

First of all... thanks a lot for doing this Pauliina. I really appreciate it. 

I created this document with extra code and some extra points to make during the lab. Feel free to add or change things in terms of code or explanation depending on what you perceive during the lab.

Again... thanks a lot!

***************

- You are going to need the following packages, make sure that you and the students load them at the beggining:

    - `library(dplyr)`  
    - `library(ggplot2)` 
    - `library(mosaic)`


- Remind students to go to my website and download the lab, they should know where to go



## Sampling from Ames, Iowa

 The purpose of this lab is to see how a given confidence interval (95% in this case)  will get us estimates that will usually capture the popuation parameters but not always. 

So while they practice how to draw samples, they will also learn how to estimate CI's and hopeully learn more about uncertainty by represnting it graphically.

<div id="boxedtext">
**Setting a seed:** Remind them to set a seed when they do thier own lab (not necesarilly  during section but later for the assignment. That way, when I run it it will look the same as when they run it). They can use whatever seed they want for their assignment,  we are going to use this one: `set.seed(9438024)` but they will use a different one
</div>

## The data

The main point here in relation to the data is that this  data-set is actually the **popuaiton**. We rarely have that luxury.



```{r}
load(url("https://raw.githubusercontent.com/GarciaRios/govt_3990/gh-pages/Labs/lab4/data/ames.RData"))
```

- After loading the data, briefly show them the data-set and variables.
- We start drawing a simple random sample of size 60 from the 
population.  (if this doesn't work, it's because they have not loaded `dplyr`)




```{r sample}
n <- 100
samp <- sample_n(ames, n)

```

- While the data-set has many housing variables, we are going to focus on the size of the house, represented by the variable `Lot.Area`.

```{r sample - Area}
samp %>% select(1:7) %>% 
  slice(1:15)
```  
  
- At this point read Exercise 1 and give them a couple of minutes to come out with some answers... ask the class what they did but you don't have to get the exact code just get broad ideas of what to do. But hopefully they estimated means if so.. ask around what they got, since they are using different seeds the random sample will be different and their  means will vary.

1.  Describe the distribution of homes in your sample. What would you say is the
    "typical" size within your sample? Also state precisely what you interpreted 
    "typical" to mean. later

- Skip this exercise they would be able to figure it out

1.  Would you expect another classmate's distribution to be identical to yours? 
    Would you expect it to be similar? Why or why not?

## Confidence intervals

 - Go over this section it's important


Return for a moment to the question that first motivated this lab: based on 
this sample, what can we infer about the population? Based only on this single 
sample, the best estimate of the average living area of houses sold in Ames 
would be the sample mean, usually denoted as $\bar{x}$ (here we're calling it 
`x_bar`). That serves as a good **point estimate** but it would be useful 
to also communicate how uncertain we are of that estimate. This uncertainty
can be quantified using a **confidence interval**.



A confidence interval for a population mean is of the following form
\[ \bar{x} + z^\star \frac{s}{\sqrt{n}} \]

- They already know this formula but show it again as a reminder
- Tell them that thy should by now be comfortable with calculating the mean and standard deviation of 
a sample in R. And we know that the sample size is 60. So the only remaining building
block is finding the appropriate critical value for a given confidence level.

- Instead of using the tables they have been  using so far, you are going to use the `qnorm` function for this task, which will give the critical value associated
with a given percentile under the normal distribution. Remember that confidence levels
and percentiles are not equivalent. For example, a 95% confidence level refers to the
middle 95% of the distribution, and the critical value associated with this area will
correspond to the 97.5th percentile.

- Show them this graph to illustrate but they should know already...


```{r eval=TRUE, echo=FALSE, fig.height=3, fig.width=5, fig.align='center'}
par(mar = c(3,0,0,0))
normTail(M = c(-1.96, 1.96), col = "#569BBD", axes = FALSE)
axis(1, at = c(-3, -1.96, 0, 1.96, 3), labels = c(NA, -1.96, 0, 1.96, NA), 
     cex.axis = 1.5)
text(x = 0, y = 0.15, "95%", cex = 1.5, col = "white")
text(x = -2.8, y = 0.05, "2.5%", cex = 1.5)
text(x = 2.8, y = 0.05, "2.5%", cex = 1.5)
graphics::arrows(-3.7, 0.25, 1.96, 0.25, col = "red", lwd = 3, code = 3)
text(x = -1.4, y = 0.275, "97.5%", cex = 1.5, col = "red")
```

We can find the critical value for a 95% confidence interval using
```{r z_star_95}
z_star_95 <- qnorm(0.975)
z_star_95
```
which is roughly equal to the value critical value 1.96 that you're likely
familiar with by now.

Let's finally calculate the confidence interval (try to explain what this line is doing or better yet ask them to explain what the line will do before they run it)

```{r ci}
samp %>%
  summarise(lower = mean(Lot.Area) - z_star_95 * (sd(Lot.Area) / sqrt(n)),
            upper = mean(Lot.Area) + z_star_95 * (sd(Lot.Area) / sqrt(n)))
```

After running that line, for teaching purposes it might be useful to also estimate the mean right there, Like this

```{r ci-w.mean}
samp %>%
  summarise(lower = mean(Lot.Area) - z_star_95 * (sd(Lot.Area) / sqrt(n)),
            mean = mean(Lot.Area),
            upper = mean(Lot.Area) + z_star_95 * (sd(Lot.Area) / sqrt(n)))
```
 -  You can see that the estimated mean obviouly falls right in the middle.. .

To recap: even though we don't know what the full population looks like, we're 95% 
confident that the true average size of houses in Ames lies between the values *lower* 
and *upper*. There are a few conditions that must be met for this interval to be valid.


- Skip 3 an 4 Go straight to calculating `mu`

1.  For the confidence interval to be valid, the sample mean must be normally 
    distributed and have standard error $s / \sqrt{n}$. What conditions must be 
    met for this to be true?

## Confidence levels

1.  What does "95% confidence" mean?

Again, in this case we have the rare luxury of knowing the true population mean since we 
have data on the entire population. Let's calculate this value so that
we can determine if our confidence intervals actually capture it. We'll store it in a data frame called `params` (short for population parameters), and name it `mu`.

```{r pop-mean}
params <- ames %>%
  summarise(mu = mean(Lot.Area))
```

- Since they will be using different seeds ask them if their `mu` was captured by their sample (Exercise 5) and skip 6
 
1.  Does your confidence interval capture the true average size of houses in 
    Ames?  does your neighbor's interval capture this value? 

1.  Each student should have gotten a slightly different confidence interval. What 
    proportion of those intervals would you expect to capture the true population 
    mean? Why?

Using R, we're going to collect many samples to learn more about how sample 
means and confidence intervals vary from one sample to another.

Here is the rough outline:

-   Obtain a random sample.
-   Calculate the sample's mean and standard deviation, and use these to calculate
and store the lower and upper bounds of the confidence intervals.
-   Repeat these steps 50 times.

We can accomplish this using the `do` function. The following lines of code takes 50
random samples of size `n` from population (and remember we defined $n = 60$ earlier),
and computes the upper and lower bounds of the confidence intervals based on these
samples.

```{r calculate-50-cis}
ci <- do(50) * ames %>%
                  sample_n(n) %>%  
  summarise(lower = mean(Lot.Area) - z_star_95 * (sd(Lot.Area) / sqrt(n)), 
            upper = mean(Lot.Area) + z_star_95 * (sd(Lot.Area) / sqrt(n)))
```

Let's view the first five intervals:

```{r first-five-intervals}
ci %>%
  slice(1:5)
```

Next we'll create a plot similar to Figure 4.8 on page 175 of [OpenIntro Statistics, 3rd
Edition](https://www.openintro.org/os). First step will be to create a new variable in 
the `ci` data frame that indicates whether the interval does or does not capture the 
true population mean. Note that capturing this value would mean the lower bound of the
confidence interval is below the value and upper bound of the confidence interval is
above the value. Remember that we create new variables using the `mutate` function.

```{r capture-mu}
ci <- ci %>%
  mutate(capture_mu = ifelse(lower < params$mu & upper > params$mu, "yes", "no"))


```


- I don't know if you use `iflelse`... it's not very intuitive so hopefully they get it... This is the first time they will use it.

The `ifelse` function is new. It takes three arguments: first is a logical statement,
second is the value we want if the logical statement yields a true result, and the
third is the value we want if the logical statement yields a false result.

 - (In this case  we use `mutate` to create a variable that would have two values: "yes" and "no". The values would be assigned using `iflese`:  if the lower bound is less than  `mu` AND  the upper bound greater than `mu`, the assigned value would be "yes", OTHERWISE it will be "no")

We now have all the information we need to create the plot, but we need to re-organize
our data a bit for easy plotting. Specifically, we need to organize the data in a new
data frame where each row represents one bound, as opposed to one interval. So this

~~~
     lower    upper capture_mu
1 1355.688 1559.545        yes
2 1413.661 1681.673        yes
3 1372.159 1631.807        yes
...
~~~

should instead look like

~~~
  ci_id ci_bounds capture_mu
1     1  1355.688        yes
2     2  1413.661        yes
3     3  1372.159        yes
4     1  1559.545        yes
5     2  1681.673        yes
6     3  1631.807        yes
...
~~~

We can accomplish this using the following:

```{r create-ci-data-for-plot}
ci_data <- data.frame(ci_id = c(1:50, 1:50),
                      ci_bounds = c(ci$lower, ci$upper),
                      capture_mu = c(ci$capture_mu, ci$capture_mu))
```

And finally we can create the plot using the following:

- if possible go over each of the lines...

```{r plot-ci}
ggplot(ci_data, aes(x = ci_bounds, y = ci_id, 
      group = ci_id, color = capture_mu)) +
  geom_point(size = 2) +  # add points at the ends, size = 2
  geom_line() +           # connect with lines
  geom_vline(xintercept = params$mu, color = "darkgray") # draw vertical line
```

 - At this point you could end or have a little discussion about the graph or the  question below


1.  What proportion of your confidence intervals include the true population mean? Is 
    this proportion exactly equal to the confidence level? If not, explain why. Make 
    sure to include your plot in your answer.

* * *

<div class="oyo">

## On your own

-   Pick a confidence level of your choosing, provided it is not 95%. What is 
    the appropriate critical value?

-   Calculate 50 confidence intervals at the confidence level you chose in the 
    previous question, and plot all intervals on one plot, and calculate the proportion 
    of intervals that include the true population mean. How does this percentage compare 
    to the confidence level selected for the intervals? Make 
    sure to include your plot in your answer.
</div>

<div id="license">
Science should be free and open source!
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
</div>